---
title: "NBA: A Look into the Three-Pointer (Part 1)" 
author: "Quang Nguyen"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE)
```

## Snippet of the Dataset
```{r, echo=FALSE}
# For Part 1, use this chunk for reading in the data with a path that identifies you.
setwd('/Users/quang/Desktop/STAT3080/project')
nba.raw = read.csv("nbastats.csv")[-c(1,10:31,41)]
head(nba.raw)
```

## Description of the Dataset - NBA Statistics from 1950 to 2017  
### Background Information
This dataset is an extensive collection of individual statistics for basketball players in the NBA from 1950 to 2017. These individual statistics were compiled from gamelogs, box-scores, and with access to the NBA developer tools (*1). The variables and relevant metrics that are included in this dataset for use in this project are included down below. Generally, the metrics that are being used measure basic statistical data concerning basketball such as points, free-throws, field-goals, attempts, minutes played, and more that can be used to find trends and correlations.  
The overall project will focus on demonstrating how the game of basketball has changed as a whole over time specifically focusing on the introduction of the three-point shot as well as how games have become faster and higher scoring. Other contexts that will be explored include the physicality of the game (using personal fouls as the indicator) and how the game has evolved to either become more offensive or defensive (based off of several metrics such as offensive rebounds versus defensive rebounds). For this portion of the project, the dataset will be used to lay the foundation for identifying correlating trends between the introduction of the three-point shot and the gradual increase of points scored through the NBA. More details will be provided about the dataset and certain methods in later sections.

### Rows and Variables
Each row within the raw dataset represents an individual player within the respective season. Each column within the raw dataset for the most part represents a unique statistic tracked by basketball analysts to help generate performance metrics of how well individual basketball players play. Of course, there are other factors that contribute to how effective a player is in the game, but these numbers provide a general overview of the performance of a player. Each column variable is described below (*2):  
"Year" - Season (recorded as the year in which the season ended so for 1949-1950 season, 1950 is reported)
"Player" - Name of the player  
"Pos" - Position of the player (could have more than one position)  
"Age" - Age of the player during that season  
"Tm" - Team that the player is on  
"G" - Total amount of games that the player played in  
"GS" - Total amount of games that the player started  
"MP" - Total amount of minutes that the player played throughout the whole season
"FG" -  Total field goals made
"FGA" - Total field goals attempted  
"FG." - Field goal percentage  
"X3P" - Total three-pointers made
"X3PA" - Total three-pointers attempted  
"X3P." - Three-point shot percentage  
"X2P" - Total two-pointers made  
"X2PA" - Total two-pointers attempted    
"X2P." - Two-point shot percentage  
"FT" - Total free throw shots made 
"FTA" - Total free throw shots attempted  
"FT." - Free throw percentage  
"ORB" - Total amount of offensive rebounds  
"DRB" - Total amount of defensive rebounds  
"TRB" - Total amount of rebounds  
"AST" - Total amount of assists  
"STL" - Total amount of steals  
"BLK" - Total amount of blocks  
"TOV" - Total amount of turnovers  
"PF" - Total amount of personal fouls  
"PTS" - Total amount of points  

### Collection of the Data
There are inconsistencies within the collection of data that exist that will be discussed in a later section. 
The data when it was collected was considered a population and contains statistics from all NBA players from 1950 to 2017. The time range chosen begins when the teams from the Basketball Association of America and the National Basketball League consolidated into the National Basketball Association prior to the 1950 season, although the NBA recognizes 1947 as the inaugural season (*3). The time range ends when the data was collected and this potential issue will be raised in a later section as well. Regarding how the data was specifically collected, the NBA's methods of maintaining statistical records have largely remained the same despite the advancement of technology. Teams of people are hired specifically to watch the basketball game and make records that can be uploaded to their database. These people are consistent in their approach as other members of the team are constantly cross-referencing and ensuring that the data is accurate. As a result, although certain metrics have been added and made more advanced which results in some null values for the earlier years, the basic information needed has been thoroughly collected in this population.  
For the purpose of my project, not all statistical recordings will be used. In the pre-subsetted dataframe, there were variables that were advanced basketball metrics such as the player efficiency rating and the offensive box plus/minus. If needed, I will calculate them based on the basic statistics that remain in the subsetted dataframe by using established formulas accepted by professional analysts. For the purpose of this investigation however, all that is needed is identifying information such as year, name, and team as well as the statistic counts for individual contributions throughout the duration of their season. Within the dataset, there are some NA values but these can also be of use to this investigation as rule changes were introduced that led to the creation of new statistic measures.

### Potential Issues
This dataset is fairly complete and exhaustive including metrics that will not be needed for this project. However, there is some missing data from the early stages of the NBA when data being stored was either not electronic, full records were not kept, or certain metrics had yet to be introduced. This can be combatted by taking a smaller sample from the population with conditions that ensure for a more complete representation. Some data is missing not because of recording errors but because of rule changes such as the addition of a three-point shot. Another minor issue is that this is not a fully updated dataset as it does not include data from the 2018 season that ended in June however, missing one year will not be a major detraction from the validity of the dataset. Overall, the dataset is comprehensive and will provide a good foundation for further analysis.

## Numerical Representation
### Five-Number Summary
```{r fig.width=7, fig.height=3}
library(data.table) # (*5)
byYear = setDT(nba.raw[,c(1,6,8:10,12:13,15:16,18:19,21:29)])[,lapply(.SD, sum),
                                                              by=Year]
summary(byYear[,c(3:20)])

```
Here, I present the data using the five number summary considering the totals for each year. This provides a more accurate representation of how the NBA has changed as the ranges of each variable (that are not percentages) are shown. In order to this, I first summed up every metric by year and then grouped them together using the data table function. This summary also details provides a clear picture of the mean versus the median for comparison which will be beneficial for further exploratory analysis.


## Graphical Representations
### Total Points Graph
```{r fig.width=7, fig.height=4, message=FALSE, warning=FALSE, results='hide'}
install.packages("ggplot2", repos = "http://cran.us.r-project.org")
library(ggplot2)
year.list = split(nba.raw, nba.raw$Year)

totpoints = sapply(year.list, function(x) sum(x$PTS))
threepointattempts = sapply(year.list, function(x) sum(x$X3PA))
dfnba = data.frame(Year=as.numeric(names(totpoints)),Points=totpoints,
                   ThreePointAttempts=threepointattempts)
ggplot(dfnba, aes(x=Year,y=Points)) + geom_point(aes(colour=Points)) +
  theme(axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.5)) +
  geom_smooth(method="lm", se=F) + scale_x_continuous(breaks = seq(1950, 2020, 5)) +
  scale_y_continuous(breaks = seq(0, 300000, 20000)) +
  labs(title="Total Points Scored by Year", y="Number of Points") +
  theme(axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.5),
        plot.title = element_text(hjust=0.5), panel.grid.minor.x = element_blank())
```
This graphical representation implicitly demonstrates how the NBA has grown. As the initial graphical representation of the data, this serves as a plot showing the total number of points that have been scored by all players throughout the years. For further exploration, certain conditions would need to be placed on the data being passed in such as how many minutes were played or the number of games during that season. That would provide a more accurate model of how the pace of the game has changed. Looking at the graph, there are certain outliers which are caused either by lockouts, player strikes, and other extraneous situations (*6). Additionally, schedules were made longer as more teams joined the league leading to more opportunities to score for an additional number of players. 

### Three-Point Attempts Graph
```{r fig.width=7, fig.height=3, warning=FALSE}
ggplot(dfnba, aes(x=Year,y=ThreePointAttempts)) +
  geom_point(aes(colour=ThreePointAttempts)) +
  theme(axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.5)) +
  geom_smooth(method="lm", se=F) + scale_x_continuous(breaks = seq(1950, 2020, 5)) +
  scale_y_continuous(breaks = seq(0, 80000, 10000)) +
  labs(title="Total Three Point Attempts Per Year Since Introduction in 1980",
       y="Number of Attempts") +
  theme(axis.text.x = element_text(angle = 90,hjust=0.95,vjust=0.5), 
        plot.title = element_text(hjust=0.5), panel.grid.minor.x = element_blank())
```
Since the three-point shots introduction into the NBA in the 1980 season, it has quickly grown in popularity among players. Today, some players have made the shot their specialty such as 2x league wide most-valuable player Steph Curry. From initial observations, it can be seen that the positive linear trend of the number three-point attempts follows the increase of points scored the graphical representation shown first.

## Initial Conclusions and Further Exploration
The numerical summaries and graphical representations provided in this portion of the investigation lays the foundation for deeper dives into the statistical world of basketball. General trends across most of the metrics provided in the raw dataset have seen large increases and the positive linear trend from my graphical representation leads me to make the initial conclusion that the game has become faster lending to more score opportunities, the size of the league has grown, and the introduction of the three-point shot has generated additional points per play. Additionally, the structure and how the game is played has almost certainly changed. The large ranges that exist in each of the five-number summaries are key indicators that rather than the players becoming better, stronger, and faster; there are also more opportunities for them to accumulate these statistical metrics. However, also using the numerical summary, these initial conclusion will be challenged as I explore further since the large range of total minutes played by all players does create some uncertainty as to whether the increase in points was simply caused by more players playing more games. For the future, I will focus on understanding whether the game has become less physical as many broadcasters have stated (*4), whether there is actual causality between the three-point shot and total points scored, and shifts between offensive/defensive mindsets throughout the years.

## References - Citations In-Text Marked by (*Reference Number)

1. Dataset retrieved from Kaggle and was mined from Basketball-Reference (*1):  
<https://www.kaggle.com/drgilermo/nba-players-stats/home>  
<https://www.basketball-reference.com/>
2. Information about basketball specific metrics (*2):  
<https://www.basketball-reference.com/about/glossary.html>
3. Information about NBA rule changes throughout history (*3):  
<http://www.nba.com/analysis/rules_history.html>
4. Article from Jabari Davis concerning physicality of the NBA (*4):  
<http://www.basketballinsiders.com/is-the-nba-becoming-too-soft/>
5. Referenced for help with data tables (*5):  
<https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html>
6. Article from CNN archives about lockouts and strikes in sports (*6):  
<https://www.cnn.com/2013/09/03/us/pro-sports-lockouts-and-strikes-fast-facts/index.html>
